Dispersion of short waves riding on longer waves

### 2024-05-09 paper/paper.tex
WIP analytical solutions

\section*{Plain Language Summary}
The ocean surface is typically covered by waves of many wavelengths.
Longer waves bunch up and dilate the shorter waves, and this process is
important to consider for both the measurements and the forecasting of ocean
waves.
Although there have been analytical and numerical solutions for how the short
waves should change in the presence of longer waves, they have been simplified
and incomplete.
This paper revisits and extends these solutions to propose a more accurate
description of how the short waves should change in the presence of longer
waves.
\section{Introduction}

Short ocean surface waves are modulated by longer swell.
Following the result by \citeA{unna1941white,unna1942waves,unna1947sea},
\citeA{longuet1960changes} introduced an analytical solution for hydrodynamic
modulation of short waves by an underlying longer wave:

\begin{equation}
\label{eq:tau}
\widetilde{k} = k (1 + a_L k_L)
\end{equation}
where $\widetilde{k}$ and $k$ are the modulated and unmodulated wavenumbers of
short waves, respectively, and $a_L$ and $k_L$ are amplitude and wavenumber
of the long wave.

This result has been influential on the development of remote sensing of ocean
surface waves and currents \cite{keller1975microwave,hara1994hydrodynamic}
and has been used even recently \cite{peureux2021unsteady} to simulate
unsteady growth of short-waves due to longer waves.

@@

Equations (\ref{eq:dispersion}) and (\ref{eq:wave_crests}) only describe the
evolution of the wavenumber.
To describe the evolution of the wave amplitude, we use the conservation of wave
action:

@@

As the short waves ride on the surface of the long wave, defined by the elevation
$\eta_L = a_L cos{\psi}$, where $\psi = k_L x - \omega_L t$ is the long-wave phase,
they move in an accelerated reference frame and will thus experience effective
gravitational acceleration that changes in both space and time.
Inserting equation (\ref{eq:dispersion}) into (\ref{eq:wave_action}) we obtain:

\begin{equation}
\label{eq:wavenumber}
\dfrac{\partial k}{\partial t}
+ \left(C_g + U\right) \dfrac{\partial k}{\partial x}
+ k \dfrac{\partial U}{\partial x}
+ \dfrac{1}{2} \sqrt{\dfrac{k}{g}} \dfrac{\partial g}{\partial x}
= 0
\end{equation}

Similarly, we expand the equation \ref{eq:wave_action} to obtain:

\begin{equation}
\label{eq:wave_action2}
\dfrac{\partial N}{\partial t}
+ \left(C_g + U\right) \dfrac{\partial N}{\partial x}
+ N \dfrac{\partial U}{\partial x}
+ N \dfrac{\partial C_g}{\partial x}
= 0
\end{equation}

\section{Analytical solutions}

\subsection{Derivation of the original L-HS solution}

\begin{equation}
\label{eq:wave_crests4}
\dfrac{d k}{k}
= - \dfrac{\partial U}{\partial x} dt
\end{equation}
which states that the relative contraction of short waves is largest where the
convergence of the ambient current $U$ is the largest.
For short waves riding on longer waves, the mean ambient current is the
horizontal near-surface orbital velocity of the long wave.
Rather than at the long-wave surface, L-HS took the ambient current to be the
long-wave orbital velocity at the mean water level, $z = 0$:

\begin{equation}
\label{eq:U_L1}
U = a_L \omega_L \cos{\psi}
\end{equation}
where $\psi$ is the long-wave phase such that the crest is located
at $\psi = 0$.
Insert (\ref{eq:U_L1}) into (\ref{eq:wave_crests4}) to obtain:

\begin{equation}
\label{eq:wave_crests5}
\dfrac{d k}{k}
= a_L k_L \omega_L \sin{\psi} dt
\end{equation}
To calculate the long-wave modulated short-wave wavenumber, integrate
(\ref{eq:wave_crests5}) from the initial time at which the short-waves
are unperturbed (at $z=0$, or $\psi=\pi/2$), to time $t$:

\begin{equation}
\label{eq:wave_crests6}
\int_k^{\ \widetilde k(t)} \dfrac{d k'}{k'}
= \int_0^t a_L k_L \omega_L \sin{\psi} dt'
\end{equation}
which yields:

\begin{equation}
\label{eq:k_short1}
\ \widetilde k(\psi) = k e^{a_L k_L \cos{\psi}}
\end{equation}
Approximating (\ref{eq:k_short1}) to the first-order Taylor expansion,
we obtain the original L-HS solution:

\begin{equation}
\label{eq:k_short2}
\ \widetilde k(\psi) = k (1 + a_L k_L \cos{\psi})
\end{equation}
The intuitive interpretation of (\ref{eq:k_short2}) is that as the short waves
contract on the front face of the long wave, they reach maximum contraction
(highest $\ \widetilde k$) at the crest of the long wave.
As they pass to the rear face of the long wave, they get elongated and reach
minimum contraction (lowest $\ \widetilde k$) in the trough of the long wave.

\subsection{Extending the L-HS solution}

L-HS made two approximations in their solution:
First, they evaluated the advective velocity of short waves at $z = 0$
rather than the surface, $z = \eta_L$, where $\eta_L$ is the long-wave
elevation; and second, they evaluated the solution to the first order
of the Taylor expansion series.
Both of these approximations cause an underestimate of the short-wave
modulation magnitude.

To extend the L-HS solution, evaluate the advective velocity at the surface
of the long-wave:

\begin{equation}
\label{eq:U_L2}
U = U_L(z=\eta) = a_L \omega_L e^{a_L k_L \cos{\psi}} \cos{\psi}
\end{equation}
and insert it into \ref{eq:wave_crests4} to obtain:

\begin{equation}
\label{eq:wave_crests7}
\dfrac{dk}{k} = a_L k_L \omega_L e^{a_L k_L \cos{\psi}} \sin{\psi} dt
\end{equation}
Integrate (\ref{eq:wave_crests7}) to the same limits as we did with
(\ref{eq:wave_crests6}) to obtain:

\begin{equation}
\label{eq:k_short3}
\widetilde{k}(\psi) = k e^{a_L k_L \cos{\psi} e^{a_L k_L \cos{\psi}}}
\end{equation}

We compare the solutions from (\ref{eq:k_short1}), (\ref{eq:k_short2}),
and (\ref{eq:k_short3}) for a moderately steep long wave with steepness
$a_L k_L = 0.2$ in Fig. (\ref{fig:wavenumber_modulation}).
The original L-HS solution is a cosine wave with an amplitude equal
to the long-wave steepness.
The other two solutions both introduce an increase in wave modulation
on the crest and a decrease in the trough.
At the crest, the relative modulation predicted by the exact L-HS solution
is 10.7\% larger than that predicted by the 1st order approximate L-HS
solution, while the solution that evaluates the long-wave orbital velocity
at the water surface rather than $z = 0$ is 38.4\% larger than that of the
1st order approximate L-HS solution.
Also, since the less-approximate solutions decrease the modulation in the
troughs and increase in on the crests, the average wavenumber modulation
is no longer 0, but 1\% in the case of the exact L-HS solution and 3.1\% in the
case of the extended solution.

\begin{figure}[h]
\label{fig:wavenumber_modulation}
\centering
\includegraphics[width=\textwidth]{figures/fig_wavenumber_modulation_by_phase.png}
\caption{
  Comparison of relative short-wave modulation as function of the
  long-wave phase, for a moderately steep long wave of $a_L k_L = 0.2$.
  The dotted line shows the 1st-order approximated solution by L-HS,
  the dashed line shows the exact solution by L-HS,
  and the solid line shows the extended solution introduced in this paper.
}
\end{figure}

\section{Apparent gravity}

@@

\section{Lagrangian averaged short-wave properties}

@@

\section{Numerical solutions}

@@

\acknowledgments
We thank anonymous reviewers for their help improving this paper.

>>>>

\section*{Plain Language Summary}
The ocean surface is typically covered by waves of many wavelengths.
Longer waves bunch up and dilate the shorter waves, and this process is
important to consider for both the measurements and the forecasting of ocean
waves.
Although there have been analytical and numerical solutions for how the short
waves should change in the presence of longer waves, they have been simplified
and incomplete.
This paper revisits and extends these solutions to propose a more accurate
description of how the short waves should change in the presence of longer
waves.

\section{Introduction}

Short ocean surface waves are modulated by longer swell.
Following the result by \citeA{unna1941white,unna1942waves,unna1947sea},
\citeA{longuet1960changes} introduced an analytical solution for hydrodynamic
modulation of short waves at the crest of an underlying longer wave:

\begin{equation}
\label{eq:tau}
\widetilde{k} = k (1 + a_L k_L)
\end{equation}
where $\widetilde{k}$ and $k$ are the modulated and unmodulated wavenumbers of
short waves, respectively, and $a_L$ and $k_L$ are amplitude and wavenumber
of the long wave, respectively.
This result has been influential on the development of remote sensing of ocean
surface waves and currents \cite{keller1975microwave,hara1994hydrodynamic}
and has been used even recently \cite{peureux2021unsteady} to simulate
unsteady growth of short-waves due to longer waves.

TODO extend with background.

@@

Equations (\ref{eq:dispersion}) and (\ref{eq:wave_crests}) only describe the
evolution of the wavenumber.
To describe the evolution of the wave amplitude, we use the conservation of wave
action in absence of sources (e.g. due to wind) and sinks (e.g. due to whitecapping):

@@

As the short waves ride on the surface of the long wave,
they move in an accelerated reference frame and will thus experience effective
gravitational acceleration that changes in both space and time.
Inserting Eq. (\ref{eq:dispersion}) into (\ref{eq:wave_action}) we obtain:

\begin{equation}
\label{eq:wavenumber}
\dfrac{\partial k}{\partial t}
+ \left(C_g + U\right) \dfrac{\partial k}{\partial x}
+ k \dfrac{\partial U}{\partial x}
+ \dfrac{1}{2} \sqrt{\dfrac{k}{g}} \dfrac{\partial g}{\partial x}
= 0
\end{equation}
From left to right, the terms in this equation represent the change in wavenumber
due to the propagation and advection by ambient current $U$, the convergence of
the ambient current, and the spatial inhomogeneity of the gravitational acceleration.
In a non-accelerated reference frame (i.e. in absence of longer waves), the last
term is zero.
However, if gravitational acceleration varies in space, this term is necessary
to conserve the wave crests.

Similar to Eq. \ref{eq:wavenumber}, we expand Eq. \ref{eq:wave_action} to obtain:

\begin{equation}
\label{eq:wave_action2}
\dfrac{\partial N}{\partial t}
+ \left(C_g + U\right) \dfrac{\partial N}{\partial x}
+ N \dfrac{\partial U}{\partial x}
+ N \dfrac{\partial C_g}{\partial x}
= 0
\end{equation}
This equation is similar to Eq. \ref{eq:wavenumber}, except for the last term
which here represents the inhomogeneity of the group speed.
As \citeA{peureux2021unsteady} explained, this term is responsible for unstable
growth of wave action in infinite long-wave trains.
Eqs. \ref{eq:wavenumber} and \ref{eq:wave_action2} will be the governing equations
that we solve numerically in section \ref{section:numerical_solutions}.

For the ambient forcing, we consider a single, monochromatic long wave
defined by the elevation $\eta_L = a_L cos{\psi}$,
where $a_L$ is the long-wave amplitude, $\psi = k_L x - \omega_L t$ is the
long-wave phase, and $\omega_L$ its angular frequency.
At this time we define the long-wave steepness $\varepsilon_L = a_L k_L$ which
we will use thoughout the rest of the paper as a scaling parameter.
The long wave induces horizontal and vertical orbital velocities whose values
at the surface $\eta_L$ are:

\begin{equation}
\label{eq:U_L}
U = a_L \omega_L e^{\varepsilon_L \cos{\psi}} \cos{\psi}
\end{equation}

\begin{equation}
\label{eq:w_L}
W = a_L \omega_L e^{\varepsilon_L \cos{\psi}} \sin{\psi}
\end{equation}

Eqs. \ref{eq:dispersion}-\ref{eq:w_L} is the full system of equations that we
exploit in the remainder of this paper.

\section{Analytical solutions}
\label{section:analytical_solutions}

\subsection{Wavenumber and amplitude modulation}

Solving analytically for the wavenumber requires dropping spatial gradients of
$k$ in Eq. \ref{eq:wavenumber} and assuming homogeneous $g$:

\begin{equation}
\label{eq:wavenumber2}
\dfrac{\partial k}{\partial t}
= - k \dfrac{\partial U}{\partial x}
\end{equation}
Although \citeA{peureux2021unsteady} pointed out that the solution by L-HS
requires assuming homogeneous group speed of the short-wave field, in fact,
it also requires assuming no horizontal propagation and advection of the short
waves.
Combining Eqs. \ref{eq:U_L} and \ref{eq:wavenumber2}, and integrating it in time
yields:

\begin{equation}
\label{eq:k_short_exact}
\widetilde{k}(\psi) = k e^{\varepsilon_L \cos{\psi} e^{\varepsilon_L \cos{\psi}}}
\end{equation}

Notice that the Taylor expansion of Eq. \ref{eq:k_short_exact} to the first order
recovers the original solution by L-HS:

\begin{equation}
\label{eq:k_short_lhs}
\widetilde{k}(\psi) = k (1 + \varepsilon_L \cos{\psi})
\end{equation}

The L-HS solution requires two approximations in their solution:
First, evaluating the long-wave velocity at $z = 0$
rather than $z = \eta_L$; and second, evaluating the solution to the first order
of the Taylor expansion series.
Both of these approximations cause an underestimate of the short-wave
modulation magnitude (Fig. \ref{fig:wavenumber_modulation}).
The original L-HS solution is a cosine wave with an amplitude equal
to the long-wave steepness.
The other two solutions both introduce an increase in wave modulation
on the crest and a decrease in the trough.
At the crest, the relative modulation predicted by the exact L-HS solution
is 10.7\% larger than that predicted by the 1st order approximate L-HS
solution, while the solution that evaluates the long-wave orbital velocity
at the water surface rather than $z = 0$ is 38.4\% larger than that of the
1st order approximate L-HS solution.

\begin{figure}[h]
\label{fig:wavenumber_modulation}
\centering
\includegraphics[width=\textwidth]{figures/fig_wavenumber_modulation_by_phase.png}
\caption{
  Comparison of relative short-wave modulation as function of the
  long-wave phase, for a moderately steep long wave of $a_L k_L = 0.1$.
  The dotted line shows the 1st-order approximated solution by L-HS,
  the dashed line shows the exact solution by L-HS,
  and the solid line shows the extended solution introduced in this paper.
}
\end{figure}

\subsection{Apparent gravity}

@@

\subsection{Lagrangian averaged short-wave properties}

@@

\section{Numerical solutions}
\label{section:numerical_solutions}

\section{Scale analysis}
\label{section:scale_analysis}

Now that we showed that the analytical solution is consistent with full numerical
solutions for $\varepsilon_L < 0.25$, we can use analytical solutions for scale
analysis to quantify the relative contribution of tendencies in the full equation set.

TODO

@@

\acknowledgments
Thanks to Peisen Tan, Nathan Laxague, and Fabrice Ardhuin for providing helpful
feedback on the early drafts of this work.
